Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2016-07-29 paint-contour-tree.tex
section 8 fixes

\subsection{Shrubs, tall paths, and short paths}
\label{sec:shrubs}

@@

\begin{lemma}
\label{lem:adjacent}
 Let $p$ be any path in $\pmax(T)$ and let $\{q_1, \dots q_k\}$ be the children of $p$.
Then $|H_{\ell(p)}| + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
\end{lemma}

\begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
Consider $v \in \bigcup_{i} H_i$ that lies below $b(p)$ in $T$. 
Note that such a vertex has only one of its two colors in $L(b(p))$.  
Since the colors tracked by $H_i$ and $H_j$ for $i\neq j$ are disjoint, 
such a vertex can appear in only one of the $H_i$'s.
On the other hand, a vertex $u\in p$ can appear in more than one $H_i$, 
but since any vertex has exactly two colors it can appear in at most two 
such heaps.  Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
\ignore{
 First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
 Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.

 Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e.\ $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
 In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
 as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
 Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
 Putting these together gives
 \[
  \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
 \]
}
\end{proof}

@@

Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
given by $\pmax(T)$. We now describe a partitioning of the unsupported paths into a forest of shrubs. Consider $p \in \cL \setminus \cL'$,
and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
Insert an edge from $p$ to $q$, so $q$ is the parent of $p$. This construction leads
to the \emph{shrub forest} of $\cL \setminus \cL'$, denoted $\cF$, where all the roots are supported short paths, 
and the remaining nodes are the unsupported short paths.

So far we know that tall paths and supported short paths can be ``paid for.''  We will also be able to pay for 
unsupported paths, though doing so is much trickier.  The following is a key property of unsupported paths which 
we will use multiple times. It is worth noting that the proof of this lemma is where the construction of $\pmax(T)$ enters the picture.

@@

\newcommand{\buff}{\mathcal{B}}
\newcommand{\loss}{\lambda}
\subsection{Pruning Away Buffered Sub-shrubs}\label{sec:buffSub}
Let $\cU$ be a shrub in $\cF$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.

@@

%We require the following lemma so that we can charge to the root in a buffered subtree.
%
% \begin{lemma}
%  Let $p$ be any node in a buffered sub-shrub $\buff$, and let $d$ be any descendant node of $p$ in $\buff$.  Then for any $v \in d$, $|H_v| = O(|H_p|)$.
% \end{lemma}
% \begin{proof}
%  Missing.  \Lem{cu-root} will be rephrased to handle this slightly more general case (i.e. here $p$ may not be the root of $\cU$). Note the proof will not 
%  have to be changed as \Lem{geometric} is the proof which already implies general case.
% \end{proof}

@@

Let $p$ and $q$ be two distinct paths in $\cF$.  Note that while it is possible that $TSC(p)\cap TSC(q)\neq \emptyset$, 
it holds that $\loss(p)\cap \loss(q) = \emptyset$.  Thus the above lemma implies that the total heap costs due to vertices in buffered sub-shrubs can be charged to those in tall paths.

\begin{corollary}
 Let $B= \{\buff_1, \ldots, \buff_k\}$ be any collection of disjoint buffered sub-shrubs from $\cF$. Then, 
 $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$
\end{corollary}

@@

\begin{definition}
 Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
 We compute a pruning of $\cU$ (see \Def{prune}), by iteratively removing all buffered sub-shrubs (see \Def{buff}), where buffered sub-shrubs whose roots are earlier in the total ordering are pruned first. We use $\pruned$ to denote the resulting sub-shrub, which we refer to as the \emph{buffer-pruned} sub-shrub of $\cU$.
\end{definition}

@@

\subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}
\textcolor{red}{\huge Ben will update this subsection to be consistent with new changes}

We split the summation into tall, supported short, and unsupported short paths.
\begin{eqnarray*} 
\sum_{p \in \pmax(T)} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
\end{eqnarray*}
The last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$, by \Lem{pathBounds}.
The second term can be bounded by $O(\sum_{p \in \cL'} |H_p| \log |H_p|)$, again by \Lem{pathBounds}.
The following claim shows that this in turn is asymptotically bounded by the last term.

\begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
\end{claim}

\begin{proof} 
Pick $p \in \cL'$. 
As we traverse the tall support chain of $p$, there
are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
Since all the vertices lie in tall paths,
we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.

The first term (unsupported short paths) can be charged to the second term (supported short paths).
This is where the critical \Lem{cu-root} plays a role.

\begin{claim} \label{clm:first} $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
\end{claim}

\begin{proof} Let $\cU$ denote a connected component of the shrub forest.
We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\cU$, and furthermore, $\sum_{p \in \cU} |p| = O(|H_r|)$. 
We have $\sum_{p \in \cU} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \cU} |p|) = O(|H_r|\log |H_r|)$.
We sum this over all $\cU$ in the shrub forest, and note that roots in the shrub forest are supported short paths.
\end{proof}

>>>>

\subsection{Tall and short paths}
\label{sec:shrubs}

@@

% \begin{lemma}
% \label{lem:adjacent}
%  Let $p$ be any path in $\pmax(T)$ and let $\{q_1, \dots q_k\}$ be the children of $p$.
% Then $|H_{\ell(p)}| + \sum_{i=1}^k |H_{q_i}| \leq |H_p|+2|p|$. 
% \end{lemma}
% 
% \begin{proof} For convenience, denote $H_i = H_{q_i}$ and $H_0 = H_{\ell(p)}$.
% Consider $v \in \bigcup_{i} H_i$ that lies below $b(p)$ in $T$. 
% Note that such a vertex has only one of its two colors in $L(b(p))$.  
% Since the colors tracked by $H_i$ and $H_j$ for $i\neq j$ are disjoint, 
% such a vertex can appear in only one of the $H_i$'s.
% On the other hand, a vertex $u\in p$ can appear in more than one $H_i$, 
% but since any vertex has exactly two colors it can appear in at most two 
% such heaps.  Hence, $\sum_i |H_i| \leq |H_p| + 2|p|$.
% \ignore{
%  First observe that for all $i$, $r_{i}$ lies above $r_p$ on some root to leaf path. 
%  Let $H_i'$ be the subset of $H_i$ that lies below $r_p$.  Observe that $H_i'\subseteq H_p$, as $L(T_{r_i})\subseteq L(T_{r_p})$.
%   
%  Now for any $i\neq j$, $H_i \cap H_j = \emptyset$, as $r_{i}$ and $r_{j}$ are not on the same root to leaf path (i.e.\ $L(T_{r_i}) \cap L(T_{r_j}) = \emptyset$).
%  In particular, $(H_i\setminus H_i') \cap (H_j \setminus H_j') = \emptyset$, and so $\sum_{i=0}^k |H_i\setminus H_i'| \leq |p|$ 
%  as $H_i\setminus H_i'$ is precisely the subset of $H_i$ that lies on $p$.
%  Moreover, for any $i\neq j$, $ H_i' \cap H_j' = \emptyset$, and so since $H_i'\subseteq H_p$ we have $\sum_{i=0}^k |H_i'|\leq |H_p|$.
%  Putting these together gives
%  \[
%   \sum_{i=0}^k |H_i| = \sum_{i=0}^k |H_i'| + \sum_{i=0}^k |H_i\setminus H_i'| \leq |H_p|+|p|.
%  \]
% }
% \end{proof}

@@

Let $\cL$ be the set of short paths, $\cL'$ be the set of supported short paths, and $\cH$ be the set of tall paths
given by $\pmax(T)$ (notation which we be used for the remainder of this section).

\begin{claim} \label{clm:above} 
$\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
\end{claim}
\begin{proof} 
Pick $p \in \cL'$. 
As we traverse the tall support chain of $p$, there
are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
$|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
Since all the vertices lie in tall paths,
we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
%and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.

\begin{corollary}\label{cor:supported}
  $\sum_{p \in \cL'} \sum_{v \in p} \log |H_v|= O(\sum_{p\in \pmax} |p|\log |p|)$
\end{corollary}
\begin{proof}
 By applying \Lem{pathBounds}, \Clm{above}, and then \Lem{pathBounds} again, we have the following.
 \begin{align*}
  \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| 
  =  O\left(\sum_{p \in \cL'} |H_p| \log |H_p|\right)
  = O\left(\sum_{p \in \cH} \sum_{v \in p} \log |H_v|\right)
  = O\left(\sum_{p\in \pmax} |p|\log |p|\right)
 \end{align*}
%
% By \Lem{pathBounds}, $\sum_{p \in \cL'} \sum_{v \in p} \log |H_v|$
\end{proof}

\subsection{Unsupported shrubs}

So far we know that tall paths and supported short paths can be ``paid for.''  We will also be able to pay for 
unsupported paths, though doing so is much trickier.

Recall that $\cL$ denotes the set of short paths, $\cL'$ the set of supported short paths, and $\cH$ the set of tall paths
given by $\pmax(T)$. We now describe a partitioning of the unsupported paths into a forest of shrubs. Consider $p \in \cL \setminus \cL'$,
and traverse the chain of ancestors from $p$. Eventually, we must reach another short path $q$.
(If not, we have reached the root $r$ of $\pmax(T)$. Hence, $p$ is supported.)
Insert an edge from $p$ to $q$, so $q$ is the parent of $p$. This construction leads
to the \emph{shrub forest} of $\cL \setminus \cL'$, denoted $\cF$, where all the roots are supported short paths, 
and the remaining nodes are the unsupported short paths.

The following is a key property of unsupported paths which 
we will use multiple times. It is worth noting that the proof of this lemma is where the construction of $\pmax(T)$ enters the picture.

@@

\newcommand{\buff}{\mathcal{B}}
\newcommand{\loss}{\lambda}
\subsection{Pruning away buffered sub-shrubs}\label{sec:buffSub}
Let $\cU$ be a shrub in $\cF$.  For any node $q$ in the shrub $\cU$, let $TSC(q)$ denote the set of all vertices of $T$ from all paths in the tall support chain of $q$.
For $q$ in $\cU$ we then define the \emph{loss} to be $\loss(q) = TSC(q)\cap H_q$. 
Intuitively, if $c$ is a child node of a parent node $p$ in $\cU$, then $\loss(c)$ are the vertices ``lost'' from the heap of $c$ to its tall support chain, by the time one reaches $p$.

@@

@@

For any fixed vertex $v$ on a tall path, there are at most two unsupported short paths such that $v$ is contained in $\loss(\cdot)$. Specifically, 
$v$ is assigned exactly two colors, one from each of its child subtrees in $T$.  Pick one of these two colors, which corresponds to some leaf in $T$.  
Any two unsupported short paths which are tracking the color of this leaf, must be ancestor-descendant in $\pmax$, however such paths have disjoint tall support chains, 
and hence only one such path can count $v$ in $\loss(\cdot)$.
(Note this is the same argument as in \Clm{above}.)  Thus by summing over all tall support chains we have the following.

%Let $p$ and $q$ be two distinct paths in $\cF$.  Note that while it is possible that $TSC(p)\cap TSC(q)\neq \emptyset$, 
%it holds that $\loss(p)\cap \loss(q) = \emptyset$.  Thus the above lemma implies that the total heap costs due to vertices in buffered sub-shrubs can be charged to those in tall paths.

\begin{corollary}\label{cor:disjointBuff}
 Let $B= \{\buff_1, \ldots, \buff_k\}$ be any collection of disjoint buffered sub-shrubs from $\cF$. Then, 
 $\sum_{i=1}^k \sum_{p\in \buff_i}\sum_{v\in p} |H_v| = O(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| )$.
\end{corollary}

@@

\begin{definition}\label{def:pruningAShrub}
 Let $\cU$ be a shrub in $\cF$. Consider a reverse topological ordering (i.e.\ leaves first) of the nodes in $\cU$, and use it to define a total ordering.  
 We compute a pruning of $\cU$ (see \Def{prune}), by iteratively removing all buffered sub-shrubs (see \Def{buff}), where buffered sub-shrubs whose roots are earlier in the total ordering are pruned first. We use $\pruned$ to denote the resulting sub-shrub, which we refer to as the \emph{buffer-pruned} sub-shrub of $\cU$.
\end{definition}

@@

\subsection{Proving \Thm{runtime}} \label{sec:thm-runtime}

We split the summation into tall, supported short, and unsupported short paths.
\begin{eqnarray*} 
\sum_{p \in \pmax(T)} \sum_{v \in p} \log |H_v| & = & \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cL'} \sum_{v \in p} \log |H_v| + \sum_{p \in \cH} \sum_{v \in p} \log |H_v|
\end{eqnarray*}
By \Cor{supported} and \Lem{pathBounds}, the second and last term can be bounded by $O(\sum_{p \in \pmax(T)} |p|\log |p|)$.
The following claim applies the results of the previous two subsections to give the same bound on the first term, 
thus completing the proof of \Thm{runtime}.

% \begin{claim} \label{clm:above} $\sum_{p \in \cL'} |H_p| \log |H_p| = O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
% \end{claim}
% 
% \begin{proof} 
% Pick $p \in \cL'$. 
% As we traverse the tall support chain of $p$, there
% are at least $|H_p|/100$ vertices of $H_p$ that lie in these paths. These are encountered in
% a fixed order. Let $H'_p$ be the first $|H_p|/200$ of these vertices. When $v \in H'_p$
% is encountered, there are $|H_p|/200$ vertices of $H_p$ not yet encountered. Hence,
% $|H_v| \geq |H_p|/200$. Hence, $|H_p|\log |H_p| = O(\sum_{v \in H'_p} \log |H_v|)$.
% Since all the vertices lie in tall paths,
% we can write this as $O(\sum_{q \in \cH} \sum_{v \in H'_p \cap q} \log |H_v|)$.
% Summing over all $p$, the expression is $\sum_{q \in \cH} \sum_{p \in \cL'} \sum_{v \in H'_p \cap q} \log |H_v|$.
% %and $\sum_{p \in \cL'} |H_p|\log |H_p| = O(\sum_{p \in \cL'} \sum_{v \in H'_p} \log |H_v|)$.
% 
% Consider any $v \in H'_p$.  Let $S$ be the set of paths $\widetilde{p}\in \cL'$ such that 
% $v\in H'_{\widetilde{p}}$.  We now show $|S|\leq 2$ (i.e.\ it contains at most one path other than $p$).
% First observe that any two paths in $S$ must be unrelated (i.e.\ $S$ is an anti-chain), 
% since paths which have an ancestor-descendant relationship have disjoint tall support chains.  
% However, any vertex $v$ receives exactly one color from each of its two subtrees (in $T$), and therefore $|S|\leq 2$ since any two 
% paths which share descendant leaves in $T$ (i.e.\ their heaps are tracking the same color) must have an ancestor-descendant relationship.
% 
% In other words, any $\log |H_v|$ appears at most twice in the above triple summation.
% Hence, we can bound it by $O(\sum_{q \in \cH} \sum_{v \in q} \log |H_v|)$.
% \end{proof}

% The first term (unsupported short paths) can be charged to the second term (supported short paths).
% This is where the critical \Lem{cu-root} plays a role.

\begin{claim} \label{clm:first} 
$\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| = O(\sum_{p\in \pmax} |p|\log |p|)$.%= O(\sum_{p \in \cL'} |H_p| \log |H_p|)$.
\end{claim}

\begin{proof} 
For any component $\cU$ of the shrub forest $\cF$ of $\cL\setminus \cL'$, let $B(\cU)$ denote the collection of buffered sub-shrubs that we 
removed in the pruning process of \Def{pruningAShrub} that produced the buffer-pruned sub-shrub $\pruned$
(i.e.\ these are disjoint sub-shrubs whose union cover $\cU\setminus (\pruned)$). 
Thus we can decompose the sum as follows, 
\begin{align*}
 \sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| 
 &\leq \sum_{\cU\in \cF} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|
 = \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  \sum_{\cU\in \cF}\sum_{\buff\in B(\cU)} \sum_{p \in \buff} \sum_{v \in p} \log |H_v|\\
 &= \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  O\left(\sum_{p\in \cH}\sum_{v\in p} \log |H_v| \right)\\
 &= \sum_{\cU\in \cF}\sum_{p \in \pruned} \sum_{v \in p} \log |H_v|  +  O\left(\sum_{p\in \pmax} |p|\log |p|\right),
\end{align*}
where the last two lines follow from \Cor{disjointBuff} and \Lem{pathBounds}, respectively.

So what remains is to bound the first term.  
%We have $\sum_{p \in \cL \setminus \cL'} \sum_{v \in p} \log |H_v| \allowbreak \leq \sum_{\cU} \sum_{p \in \cU} \sum_{v \in p} \log |H_v|$.
By \Lem{cu-root}, $|H_v| = O(|H_r|)$, where $r$ is the root of $\pruned$, and furthermore, $\sum_{p \in \pruned} |p| = O(|H_r|)$. 
We have $\sum_{p \in \pruned} \sum_{v \in p} \log |H_v| = O((\log |H_r|) \sum_{p \in \pruned} |p|) = O(|H_r|\log |H_r|)$.
Note that the roots in the shrub forest are supported short paths, and hence when we sum this over all $\cU$ in the shrub forest we get, 
\[
\sum_{\cU\in \cF} O\left(|H_{r(\pruned)}|\log |H_{r(\pruned)}|\right)
=\sum_{p\in \cL'} O(|H_p|\log |H_p|) = O\left(\sum_{p\in \pmax} |p|\log |p|\right),
\]
where the last equality follows by \Clm{above} (via the same argument as in \Cor{supported}).
\end{proof}
